### 2024-09-19 outputs/paper/paper.qmd
changed title of paper

---
title: "Choosing TTC x Provides the Most Time Stability during Rush Hours"
subtitle: "Analysis on the Delay Magnitude of different TTC Services"
author: 
  - Sandy Yu
thanks: "Code and data are available at: https://github.com/Jingying-yu/central_intake_calls_and_homeless_death_count.git"
date: today
date-format: long
abstract: "For the City of Toronto, the physical well-being of its local homeless population has always been one of its primary focuses. Numerous social efforts, funded by the City, were made in attempts to reduce the death rate of the local homeless population. In order to visualize the magnitude of social impact enforced by the City of Toronto, analyses were performed to investigate the correlation between the efforts made by the staff of Toronto’s Central Intake Line and the monthly homeless death counts between November 2020 to June 2023 using data from OpenDataToronto. Staff efforts such as the number of calls coded, number of referral to shelter, and number of information for homelessness prevention, were documented and compared against the monthly homeless death counts. An inverse correlation is found between the number of calls coded and the homeless death count. No discernible correlations were found between coded call types of interest and the monthly homeless death counts."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(lubridate)
library(zoo)
library(knitr)
library(ggplot2)
```

# Introduction

The City of Toronto has been consistently providing funding and staff in an effort to reduce the active homeless population within its governing district. One of the key aspects of reducing homelessness is the assignment of sheltering space. Amongst numerous city efforts, the 24/7 helpline — Toronto Central Intake Line — was created to offer referrals to emergency shelters, sleeping accommodations, and provide general information about homelessness and prevention. [@Central_Intake_Call]. Starting in 2020, detailed data about the Central Intake Line became available on the OpenDataToronto portal [@OpenDataToronto].

In this paper, we want to measure the magnitude of positive impact created by the Central Intake Line in relation to the monthly cumulative number of deaths in the homeless community. Using Central Intake Call Wrap-up Codes data [@Central_Intake_Call] from OpenDataToronto [@OpenDataToronto], three different aspects of the Central Intake Line efforts were isolated: total calls coded, referral to shelter, and homelessness prevention information provided. These three key aspects were measured against the monthly cumulative Deaths of People Experiencing Homelessness Data [@Homeless_Death_Count] documented by the Toronto Public Health [@TPH]. The analyses yielded an inverse correlation between the number of calls coded and the homeless death count, and no discernible correlations between coded call types of interest and the monthly homeless death counts.

Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-result -- Result, @sec-discussion -- Discussion, and @sec-conclusion -- Conclusion. The Data section examines all datasets and variables kept for analysis, followed by an explanation of their data cleaning processes. The Result section focuses on visualizing and presenting the correlation between variables of interest. The Discussion sections further evaluate the meaning behind the correlation presented in the previous section. Lastly, the arguments presented in all previous sections are wrapped up by the Conclusion section which summarizes the main findings in this paper.

# Data {#sec-data}

All data used in this paper are obtained through OpenDataToronto Portal [@OpenDataToronto]. Two different datasets: Central Intake Call Wrap-Up Codes Data [@Central_Intake_Call] and Deaths of People Experiencing Homelessness[@Homeless_Death_Count], are retrieved to analyze the correlation between different factors of Toronto’s Central Call Line efforts and the monthly cumulative death counts of homeless individuals in Toronto. Data was cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], lubridate [@lubridate], zoo [@citeZoo], ggplot2 [@ggplot2], and knitr [@knitr]. Detailed description of each dataset can be found in the subsections below.

## Central Intake Call Wrap-Up Codes

On the OpenDataToronto portal, there are several datasets that reflect the City’s effort to shelter the local homeless population. The Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call], stored in the Central Intake Calls Catalogue [@Central_Intake_Call], is one of the freshest and most detailed. Data was stored and published by the Shelter, Support & Housing Administration since November of 2020 and refreshes on a monthly basis. The latest refresh occurred on January 15th, 2024.

The data set provides a daily summary of the number of calls received, the number of calls classified into distinct wrap-up codes by the nature of the issue, and a count of calls under each wrap-up code. One of the example wrap-up codes in the original data set was: *Code 1A - Referral to a Sleeping/Resting Space*. The original data set includes 13 distinct wrap-up codes; only two codes, *Code 1A — Referral to a Sleeping/Resting Space* and *Code 2C — Information - Homelessness & Prevention Services*, were chosen for our analysis. Code 1A and Code 2C are best suited as measurements for positive impact the Central Intake Line could provide because they provide a count of the number of callers provided with directions or advice. Other non-suitable wrap-up codes include: *Code 1D - Declined Shelter/Resting Space*, *Code 4B - Disconnected - No Outcome*, etc.

The final dataset only includes data before July of 2023. The rationalization behind the filtering and summarizing actions will be explained in @sec-data-diff-datasets.

```{r}
#| label: tbl-monthly-calls
#| tbl-cap: Sample of Cleaned Central Intake Call Wrap-up Codes Dataset
#| echo: false

central_intake_call <- read.csv(here::here("outputs/data/cleaned_central_intake_call.csv"))

sample_central_intake_call <- head(central_intake_call)

knitr::kable(sample_central_intake_call, col.names = c("Month", "Calls Coded", "Referral to Shelter", "Homelessness Prevention Info"), align = c("l", "c", "c", "c"))
```

## Deaths of People Experiencing Homelessness

The Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] contains monthly cumulative records of homeless deaths. The dataset is published by Toronto Public Health [@TPH]. The earliest data record started in January of 2017, and the latest record ends in June of 2023.

The original dataset contains three columns: *Year of death*, *Month of death*, and *Count*. After careful inspection of raw data, any rows that containing an “unknown” value are excluded. Although this action created a source of error in future analyses, this is still a necessary step because we are plotting death counts against a timeline.

Final dataset only includes data from November 2020 to June 2023 (@tbl-death-count). The rationalization behind the filtering will be explained in @sec-data-diff-datasets.

```{r}
#| label: tbl-death-count
#| tbl-cap: Sample of Cleaned Deaths of People Experiencing Homelessness Dataset
#| echo: false

homeless_death_counts <- read.csv(here::here("outputs/data/cleaned_homeless_death_counts.csv"))

sample_homeless_death_counts <- head(homeless_death_counts, n = 3)

knitr::kable(sample_homeless_death_counts, col.names = c("Last Day of Month", "Death Count"), align = c("c", "c"))
```

## Using Data from Different Datasets {#sec-data-diff-datasets}
During the data cleaning process, code chunks are written to ensure that both the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] are filtered to include only data between 1st November, 2020 to 30st June, 2023.

The decision to only include this time period is reached by taking the common time period between the two datasets. In the latest refresh, the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] begins on 3rd November, 2020 and end on 31st December, 2023; whereas the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] begins on January of 2017 and ends on June 2023.

The ultimate purpose of taking the common time period is to ensure that we can plot variables of interest in the two datasets onto the same graph. Thus a combined dataset that kept the variables of interest in both datasets is created (@tbl-combined).
```{r}
#| label: tbl-combined
#| tbl-cap: Sample of Combined dataset
#| echo: false

combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
                          show_col_types = FALSE)

sample_combined_data <- head(combined_data, n = 3)

knitr::kable(sample_combined_data, col.names = c("Month", "Calls Coded", "Referral to Shelter", "Homelessness Prevention Info", "Death Count"), align = c("l","c", "c", "c","c"))
```

# Results {#sec-result}

Between November 2020 and June 2023, there were a total of 522 cases of homeless death recorded in Toronto. The mean monthly death count is 16.3125, with a minimum of 8 recorded cases and a maximum of 26 recorded cases per month. The overall trend of homeless death cases is downward-sloping.

## Number of Calls Coded & Homeless Death Count
The first graph (@fig-coded) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted cumulative number of calls coded per month. The value of calls coded per month, represented by the red line, is divided by 1000 for easier comparison against the death count. Original values are too big for any meaningful comparison.

The red line in the graph visualizes the number of calls coded per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023. Throughout the time period, there were a total of 474515 calls coded, the mean monthly number of calls coded is 14828.59. The minimum number of calls coded per month was 8367, and the maximum number of calls coded was 30520. We can see a positive growth in the number of calls being coded per month over time (@fig-coded).
```{r}
#| label: fig-coded
#| fig-cap: Visualization to show correlation between number of Calls Coded (represented by the red line) and Homeless Death Counts (represented by the blue line)
#| echo: false
#| message: false

combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
                          show_col_types = FALSE)

combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
  geom_point() + geom_smooth() + 
  geom_line(aes(x = Month, y = Calls_Coded_month/1000), color = "red") +
  labs(
    title = "Calls Coded & Homeless Death Count",
    x = "Time (in Month)",
    y = "Count"
  )
#still need to add a legend or label, but I haven't yep figured out how
```

## Number of Callers Referred to Shelter & Homeless Death Count
The second graph (@fig-referred) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted cumulative number of callers referred to shelters per month. The number of callers referred to shelters per month, represented by the purple line, is divided by 100 for easier comparison against the death count. Original values are too big for any meaningful comparison.

The number of callers referred to shelters per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023 is represented by the blue line in @fig-referred. Throughout the time period, there were a total of 44760 callers referred, the mean monthly number of callers referred was 1398.75. The minimum number of callers referred per month was 456, and the maximum number of callers referred was 2192. We can observe a negative growth in the number of callers referred per month over time (@fig-referred).
```{r}
#| label: fig-referred
#| fig-cap: Visualization to show correlation between number of callers referred to shelter (represented by the purple line) and Homeless Death Counts (represented by the blue line)
#| echo: false
#| message: false

combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
                          show_col_types = FALSE)

combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
  geom_point() + geom_smooth() + 
  geom_line(aes(x = Month, y = Referral_to_Shelter_month/100), color = "purple") +
  labs(
    title = "Callers Referred to Shelter & Homeless Death",
    x = "Time (in Month)",
    y = "Count"
  )
#still need to add a legend or label, but I haven't yep figured out how
```

## Number of Callers given information on Homelessness Prevention & Homeless Death Count
The third graph (@fig-informed) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted monthly cumulative number of callers provided with information about homelessness prevention. The number of callers provided with information, represented by the dark green line, is divided by 100 for easier comparison against the death count. Original values are too big for any meaningful comparison.

The number of callers provided with information on homelessness prevention per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023 is represented by the dark green line in @fig-informed. Throughout the time period, there were a total of 58333 callers provided with information, the mean monthly number of callers provided with information was 1822.906. The minimum number of callers provided with information per month was 1110, and the maximum number of callers provided with information was 2500. We can observe an initial negative growth followed by a positive growth in the number of callers provided with information per month over time (@fig-informed).
```{r}
#| label: fig-informed
#| fig-cap: Visualization to show correlation between number of callers given information on homelessness prevention (represented by the dark green line) and Homeless Death Counts (represented by the blue line)
#| echo: false
#| message: false

combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
                          show_col_types = FALSE)

combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
  geom_point() + geom_smooth() + 
  geom_line(aes(x = Month, y = Information_Homelessness_and_Prevention_month/100), color = "darkgreen") +
  labs(
    title = "Callers given Info on Homeless Prevention vs Homeless Death",
    x = "Time (in Month)",
    y = "Count"
  )
#still need to add a legend or label, but I haven't yep figured out how
```

# Discussion {#sec-discussion}

## Correlation between number of Calls Coded & Homeless Death Counts {#sec-relation-calls}
Based on @fig-coded, there seems to be an inverse relationship between the directions of change for the homeless death count trend over time, the blue line, and the monthly cumulative calls coded represented by the red lines. When the red line moves upward, the blue line moves down. This inverse relationship indicates a negative correlation between the two variables. A negative correlation between these two variables implies that the efforts the Central Intake Line placed into coding calls has created a positive impact in reducing homeless death in Toronto.

## Correlation between number of Coded Calls of Interest & Homeless Death Counts
According to the @fig-referred and @fig-informed, both the number of callers referred to shelters (represented by the purple line in @fig-referred) and the number of callers provided with information on homelessness prevention (represented by the dark green line in @fig-informed) change in the same direction as the monthly cumulative homeless death count. Through pure visual observation, a positive correlation between homeless death count to each of the variables would be found. This relationship would imply that the Central Intake Line [@Central_Intake_Call] has actually created a negative impact in reducing the homeless death count in Toronto by referring callers to shelter and providing callers with information on homelessness prevention. However, the magnitude of fluctuation for both the purple and dark green line are very large throughout the entire time period. Especially from the beginning of 2022 to the end of 2023 for the purple line. Due to the magnitude of fluctuation for both lines, we cannot responsibly conclude any causal relationship between the two variables by pure observation of the trends. We can only say that the general trend of the purple and dark green line moves in the same direction as the homeless death count trend. Therefore, no meaningful correlation between the two variables were concludable using only the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] obtained from OpenDataToronto portal [@OpenDataToronto]. However, this result does not imply that no correlation between variables exists in the real world. It is completely plausible that correlation between the two variables can be found if more detailed data were provided, or more advanced statistical models were used for analysis.

## Errors in analysis, Limitations and Next Steps

Potential errors exist in the final datasets used for analyses. During the data cleaning process,  if rows in the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] contain “Unknown” as a Month or Year value, then these rows are filtered out of the final dataset used for relationship analyses. This filtering action would cause an inaccuracy in the total cumulation of the homeless death count; however, incoherence of dates would occur if these rows were left inside the dataset. Another potential error could occur even before I download datasets from OpenDataToronto [@OpenDataToronto]. We cannot rule out the possibility of human error during the data inputting process before the data is submitted to the OpenDataToronto portal [@OpenDataToronto].

Limitations to this analysis exist for a few reasons. Firstly, each Central Intake Line staff has their own understanding about which code they should assign their calls. We have no way of verifying whether the staff have categories their calls correctly. Secondly, the staff do not have enough time to conduct repeated follow-ups with each caller they were assigned to. We have no way of verifying what percentage of callers referred to a shelter actually got there. Lastly, we do not have enough information to sort the coded calls by demography. A subset of data that sorts the homeless death count by demography exists in the Deaths of People Experiencing Homelessness Catalogue [@Homeless_Death_Count], if more information about the demography of the callers were provided, more information about the relationship between variables of interest might be found. Further analysis on homeless death counts by demography can be conducted based on shelter data that contains demographic values.

# Conclusion {#sec-conclusion}

This paper investigated the magnitude of impact created by the Central Intake Line in relation to the monthly cumulative number of deaths in the homeless community. Variables of interest such as: calls coded, number of callers referred to shelter, number of callers provided with information on homelessness prevention, and homeless death count, were obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] stored in the OpenDataToronto portal [@OpenDataToronto]. Statistical analyses yielded a negative correlation between the number of calls coded and the homeless death count, which implies a positive impact in reducing homeless death in Toronto. Other variables, such as number of callers referred to shelter and number of callers provided with information on homelessness prevention, yielded no meaningful correlation to the homeless death count. Therefore we cannot conclude the magnitude of their contribution to the reduction of homeless death count. Overall, we can only conclude that the effort of Coding Calls have a positive impact on reducing homeless death in Toronto. Further analysis on homeless death counts by demography can be conducted based on shelter data that contains demographic values.

\newpage

# References

### 2024-09-20 outputs/paper/paper.qmd
constructed paper structure and specific topics for each sub-section

@@
 
-The City of Toronto has been consistently providing funding and staff in an effort to reduce the active homeless population within its governing district.
-One of the key aspects of reducing homelessness is the assignment of sheltering space.
-Amongst numerous city efforts, the 24/7 helpline — Toronto Central Intake Line — was created to offer referrals to emergency shelters, sleeping accommodations, and provide general information about homelessness and prevention.
-[@Central_Intake_Call].
-Starting in 2020, detailed data about the Central Intake Line became available on the OpenDataToronto portal [@OpenDataToronto].
-
-In this paper, we want to measure the magnitude of positive impact created by the Central Intake Line in relation to the monthly cumulative number of deaths in the homeless community.
-Using Central Intake Call Wrap-up Codes data [@Central_Intake_Call] from OpenDataToronto [@OpenDataToronto], three different aspects of the Central Intake Line efforts were isolated: total calls coded, referral to shelter, and homelessness prevention information provided.
-These three key aspects were measured against the monthly cumulative Deaths of People Experiencing Homelessness Data [@Homeless_Death_Count] documented by the Toronto Public Health [@TPH].
-The analyses yielded an inverse correlation between the number of calls coded and the homeless death count, and no discernible correlations between coded call types of interest and the monthly homeless death counts.
-
-Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-result -- Result, @sec-discussion -- Discussion, and @sec-conclusion -- Conclusion.
-The Data section examines all datasets and variables kept for analysis, followed by an explanation of their data cleaning processes.
-The Result section focuses on visualizing and presenting the correlation between variables of interest.
-The Discussion sections further evaluate the meaning behind the correlation presented in the previous section.
-Lastly, the arguments presented in all previous sections are wrapped up by the Conclusion section which summarizes the main findings in this paper.
-
 # Data {#sec-data}
 
-All data used in this paper are obtained through OpenDataToronto Portal [@OpenDataToronto].
-Two different datasets: Central Intake Call Wrap-Up Codes Data [@Central_Intake_Call] and Deaths of People Experiencing Homelessness[@Homeless_Death_Count], are retrieved to analyze the correlation between different factors of Toronto’s Central Call Line efforts and the monthly cumulative death counts of homeless individuals in Toronto.
-Data was cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], lubridate [@lubridate], zoo [@citeZoo], ggplot2 [@ggplot2], and knitr [@knitr].
-Detailed description of each dataset can be found in the subsections below.
-
-## Central Intake Call Wrap-Up Codes
-
-On the OpenDataToronto portal, there are several datasets that reflect the City’s effort to shelter the local homeless population.
-The Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call], stored in the Central Intake Calls Catalogue [@Central_Intake_Call], is one of the freshest and most detailed.
-Data was stored and published by the Shelter, Support & Housing Administration since November of 2020 and refreshes on a monthly basis.
-The latest refresh occurred on January 15th, 2024.
-
-The data set provides a daily summary of the number of calls received, the number of calls classified into distinct wrap-up codes by the nature of the issue, and a count of calls under each wrap-up code.
-One of the example wrap-up codes in the original data set was: *Code 1A - Referral to a Sleeping/Resting Space*.
-The original data set includes 13 distinct wrap-up codes; only two codes, *Code 1A — Referral to a Sleeping/Resting Space* and *Code 2C — Information - Homelessness & Prevention Services*, were chosen for our analysis.
-Code 1A and Code 2C are best suited as measurements for positive impact the Central Intake Line could provide because they provide a count of the number of callers provided with directions or advice.
-Other non-suitable wrap-up codes include: *Code 1D - Declined Shelter/Resting Space*, *Code 4B - Disconnected - No Outcome*, etc.
-
-The final dataset only includes data before July of 2023. The rationalization behind the filtering and summarizing actions will be explained in @sec-data-diff-datasets.
+## TTC Bus Delay
 
@@
 
-## Deaths of People Experiencing Homelessness
-
-The Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] contains monthly cumulative records of homeless deaths.
-The dataset is published by Toronto Public Health [@TPH].
-The earliest data record started in January of 2017, and the latest record ends in June of 2023.
-
-The original dataset contains three columns: *Year of death*, *Month of death*, and *Count*.
-After careful inspection of raw data, any rows that containing an “unknown” value are excluded.
-Although this action created a source of error in future analyses, this is still a necessary step because we are plotting death counts against a timeline.
-
-Final dataset only includes data from November 2020 to June 2023 (@tbl-death-count). The rationalization behind the filtering will be explained in @sec-data-diff-datasets.
+## TTC Streetcar Delay
 
@@
 
-## Using Data from Different Datasets {#sec-data-diff-datasets}
-During the data cleaning process, code chunks are written to ensure that both the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] are filtered to include only data between 1st November, 2020 to 30st June, 2023.
+## TTC Subway Delay
 
-The decision to only include this time period is reached by taking the common time period between the two datasets.
-In the latest refresh, the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] begins on 3rd November, 2020 and end on 31st December, 2023; whereas the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] begins on January of 2017 and ends on June 2023.
-
-The ultimate purpose of taking the common time period is to ensure that we can plot variables of interest in the two datasets onto the same graph.
-Thus a combined dataset that kept the variables of interest in both datasets is created (@tbl-combined).
 ```{r}
@@
 
+## TTC Delay Comparision
+
 # Results {#sec-result}
 
-Between November 2020 and June 2023, there were a total of 522 cases of homeless death recorded in Toronto.
-The mean monthly death count is 16.3125, with a minimum of 8 recorded cases and a maximum of 26 recorded cases per month.
-The overall trend of homeless death cases is downward-sloping.
+## Highest Number of Total Delayed
 
-## Number of Calls Coded & Homeless Death Count
-The first graph (@fig-coded) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted cumulative number of calls coded per month.
-The value of calls coded per month, represented by the red line, is divided by 1000 for easier comparison against the death count.
-Original values are too big for any meaningful comparison.
-
-The red line in the graph visualizes the number of calls coded per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023.
-Throughout the time period, there were a total of 474515 calls coded, the mean monthly number of calls coded is 14828.59.
-The minimum number of calls coded per month was 8367, and the maximum number of calls coded was 30520.
-We can see a positive growth in the number of calls being coded per month over time (@fig-coded).
 ```{r}
@@
 
-## Number of Callers Referred to Shelter & Homeless Death Count
-The second graph (@fig-referred) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted cumulative number of callers referred to shelters per month.
-The number of callers referred to shelters per month, represented by the purple line, is divided by 100 for easier comparison against the death count.
-Original values are too big for any meaningful comparison.
+## Most Delayed during Weekday
+Define Rush Hours
 
-The number of callers referred to shelters per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023 is represented by the blue line in @fig-referred.
-Throughout the time period, there were a total of 44760 callers referred, the mean monthly number of callers referred was 1398.75.
-The minimum number of callers referred per month was 456, and the maximum number of callers referred was 2192.
-We can observe a negative growth in the number of callers referred per month over time (@fig-referred).
+Present Results During Rush Hours & at Off-Peak Hours
 ```{r}
@@
 
-## Number of Callers given information on Homelessness Prevention & Homeless Death Count
-The third graph (@fig-informed) maps the overall trend of homeless death over time, represented by the blue line, against the adjusted monthly cumulative number of callers provided with information about homelessness prevention.
-The number of callers provided with information, represented by the dark green line, is divided by 100 for easier comparison against the death count.
-Original values are too big for any meaningful comparison.
+## Most Delayed during Weekends
 
-The number of callers provided with information on homelessness prevention per month obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] between November 2020 and June 2023 is represented by the dark green line in @fig-informed.
-Throughout the time period, there were a total of 58333 callers provided with information, the mean monthly number of callers provided with information was 1822.906.
-The minimum number of callers provided with information per month was 1110, and the maximum number of callers provided with information was 2500.
-We can observe an initial negative growth followed by a positive growth in the number of callers provided with information per month over time (@fig-informed).
 ```{r}
@@
 
+## Most Popular Delay Reasons
+- 1 table for all three services
+
+## Most Popular Delayed Vehicles
+- 1 table for all three services
+
 # Discussion {#sec-discussion}
 
-## Correlation between number of Calls Coded & Homeless Death Counts {#sec-relation-calls}
-Based on @fig-coded, there seems to be an inverse relationship between the directions of change for the homeless death count trend over time, the blue line, and the monthly cumulative calls coded represented by the red lines.
-When the red line moves upward, the blue line moves down.
-This inverse relationship indicates a negative correlation between the two variables.
-A negative correlation between these two variables implies that the efforts the Central Intake Line placed into coding calls has created a positive impact in reducing homeless death in Toronto.
+## TTC Service Prices {#sec-price}
+- one-fare
 
-## Correlation between number of Coded Calls of Interest & Homeless Death Counts
-According to the @fig-referred and @fig-informed, both the number of callers referred to shelters (represented by the purple line in @fig-referred) and the number of callers provided with information on homelessness prevention (represented by the dark green line in @fig-informed) change in the same direction as the monthly cumulative homeless death count.
-Through pure visual observation, a positive correlation between homeless death count to each of the variables would be found.
-This relationship would imply that the Central Intake Line [@Central_Intake_Call] has actually created a negative impact in reducing the homeless death count in Toronto by referring callers to shelter and providing callers with information on homelessness prevention.
-However, the magnitude of fluctuation for both the purple and dark green line are very large throughout the entire time period.
-Especially from the beginning of 2022 to the end of 2023 for the purple line.
-Due to the magnitude of fluctuation for both lines, we cannot responsibly conclude any causal relationship between the two variables by pure observation of the trends.
-We can only say that the general trend of the purple and dark green line moves in the same direction as the homeless death count trend.
-Therefore, no meaningful correlation between the two variables were concludable using only the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] obtained from OpenDataToronto portal [@OpenDataToronto].
-However, this result does not imply that no correlation between variables exists in the real world.
-It is completely plausible that correlation between the two variables can be found if more detailed data were provided, or more advanced statistical models were used for analysis.
+## Correlation between Delay and Direction of Travel
+- Financial District
+- Midtown Direction
+
+## TTC Service Amenities
+- The PATH (for financial district)
+- AC in Summer & Winter
 
@@
 
-Potential errors exist in the final datasets used for analyses.
-During the data cleaning process,  if rows in the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] contain “Unknown” as a Month or Year value, then these rows are filtered out of the final dataset used for relationship analyses.
-This filtering action would cause an inaccuracy in the total cumulation of the homeless death count; however, incoherence of dates would occur if these rows were left inside the dataset.
-Another potential error could occur even before I download datasets from OpenDataToronto [@OpenDataToronto].
-We cannot rule out the possibility of human error during the data inputting process before the data is submitted to the OpenDataToronto portal [@OpenDataToronto].
-
-Limitations to this analysis exist for a few reasons.
-Firstly, each Central Intake Line staff has their own understanding about which code they should assign their calls.
-We have no way of verifying whether the staff have categories their calls correctly.
-Secondly, the staff do not have enough time to conduct repeated follow-ups with each caller they were assigned to.
-We have no way of verifying what percentage of callers referred to a shelter actually got there.
-Lastly, we do not have enough information to sort the coded calls by demography.
-A subset of data that sorts the homeless death count by demography exists in the Deaths of People Experiencing Homelessness Catalogue [@Homeless_Death_Count], if more information about the demography of the callers were provided, more information about the relationship between variables of interest might be found.
-Further analysis on homeless death counts by demography can be conducted based on shelter data that contains demographic values.
-
 # Conclusion {#sec-conclusion}
-
-This paper investigated the magnitude of impact created by the Central Intake Line in relation to the monthly cumulative number of deaths in the homeless community.
-Variables of interest such as: calls coded, number of callers referred to shelter, number of callers provided with information on homelessness prevention, and homeless death count, were obtained from the Central Intake Call Wrap-Up Codes Dataset [@Central_Intake_Call] and the Deaths of People Experiencing Homelessness Dataset [@Homeless_Death_Count] stored in the OpenDataToronto portal [@OpenDataToronto].
-Statistical analyses yielded a negative correlation between the number of calls coded and the homeless death count, which implies a positive impact in reducing homeless death in Toronto.
-Other variables, such as number of callers referred to shelter and number of callers provided with information on homelessness prevention, yielded no meaningful correlation to the homeless death count.
-Therefore we cannot conclude the magnitude of their contribution to the reduction of homeless death count.
-Overall, we can only conclude that the effort of Coding Calls have a positive impact on reducing homeless death in Toronto.
-Further analysis on homeless death counts by demography can be conducted based on shelter data that contains demographic values.
 

### 2024-09-21 outputs/paper/paper.qmd
added visualizations to the data section of the paper

@@
   - Sandy Yu
-thanks: "Code and data are available at: https://github.com/Jingying-yu/central_intake_calls_and_homeless_death_count.git"
+thanks: "Code and data are available at: https://github.com/Jingying-yu/ttc_delay.git"
 date: today
 date-format: long
-abstract: "For the City of Toronto, the physical well-being of its local homeless population has always been one of its primary focuses.
-Numerous social efforts, funded by the City, were made in attempts to reduce the death rate of the local homeless population.
-In order to visualize the magnitude of social impact enforced by the City of Toronto, analyses were performed to investigate the correlation between the efforts made by the staff of Toronto’s Central Intake Line and the monthly homeless death counts between November 2020 to June 2023 using data from OpenDataToronto.
-Staff efforts such as the number of calls coded, number of referral to shelter, and number of information for homelessness prevention, were documented and compared against the monthly homeless death counts.
-An inverse correlation is found between the number of calls coded and the homeless death count.
-No discernible correlations were found between coded call types of interest and the monthly homeless death counts."
+abstract: ""
 format: pdf
@@
 
+Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion.
+The Data section introduces the context of the datasets and examines all datasets and variables kept for analysis.
+The Result section focuses on visualizing and presenting the data results in Data section.
+The Discussion section further evaluate the implications behind the data results presented in the previous section, expands the topic beyond pure data, and touches on any weaknesses and next steps.
+
 # Data {#sec-data}
@@
 ```{r}
-#| label: tbl-monthly-calls
-#| tbl-cap: Sample of Cleaned Central Intake Call Wrap-up Codes Dataset
+#| label: tbl-bus-delay
+#| tbl-cap: Sample of Cleaned TTC Bus Delay Dataset
 #| echo: false
 
-central_intake_call <- read.csv(here::here("outputs/data/cleaned_central_intake_call.csv"))
+bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus_delay.csv"))
 
-sample_central_intake_call <- head(central_intake_call)
+sample_bus_delay <- head(bus_delay, n = 3)
 
-knitr::kable(sample_central_intake_call, col.names = c("Month", "Calls Coded", "Referral to Shelter", "Homelessness Prevention Info"), align = c("l", "c", "c", "c"))
+knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Direction", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
 ```
@@
 ```{r}
-#| label: tbl-death-count
-#| tbl-cap: Sample of Cleaned Deaths of People Experiencing Homelessness Dataset
+#| label: tbl-streetcar-delay
+#| tbl-cap: Sample of Cleaned TTC Streetcar Delay Dataset
 #| echo: false
 
-homeless_death_counts <- read.csv(here::here("outputs/data/cleaned_homeless_death_counts.csv"))
+streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar_delay.csv"))
 
-sample_homeless_death_counts <- head(homeless_death_counts, n = 3)
+sample_streetcar_delay <- head(streetcar_delay, n = 3)
 
-knitr::kable(sample_homeless_death_counts, col.names = c("Last Day of Month", "Death Count"), align = c("c", "c"))
+knitr::kable(sample_streetcar_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Bound", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
 ```
@@
 
+### Delay Code
 ```{r}
-#| label: tbl-combined
-#| tbl-cap: Sample of Combined dataset
+#| label: tbl-subway-code
+#| tbl-cap: Sample of TTC Subway Delay Code
 #| echo: false
+#| message: false
 
-combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
-                          show_col_types = FALSE)
+subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))
 
-sample_combined_data <- head(combined_data, n = 3)
+sample_subway_delay_code <- head(subway_delay_code, n = 3)
 
-knitr::kable(sample_combined_data, col.names = c("Month", "Calls Coded", "Referral to Shelter", "Homelessness Prevention Info", "Death Count"), align = c("l","c", "c", "c","c"))
+knitr::kable(sample_subway_delay_code, col.names = c("Code", "Code Description"), align = c("c","c"))
+```
+
+### Delay Data
+```{r}
+#| label: tbl-subway-delay
+#| tbl-cap: Sample of Cleaned TTC Subway Delay Dataset
+#| echo: false
+#| message: false
+
+subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway_delay.csv"))
+
+sample_subway_delay <- head(subway_delay, n = 3)
+
+knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Location", "Delay Code", "Min Delayed", "Min Gap", "Bound", "Line", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
 ```

### 2024-09-21 outputs/paper/paper.qmd
created new tbls and other visualizations for data & result sections

@@
 #| message: false
-
+# load librarys and datasets
 library(tidyverse)
@@
 library(ggplot2)
+
+bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus_delay.csv"))
+streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar_delay.csv"))
+subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway_delay.csv"))
+subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))
+
 ```
@@
 
-bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus_delay.csv"))
-
 sample_bus_delay <- head(bus_delay, n = 3)
@@
 #| echo: false
-
-streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar_delay.csv"))
 
@@
 
-subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))
-
 sample_subway_delay_code <- head(subway_delay_code, n = 3)
@@
 #| message: false
-
-subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway_delay.csv"))
 
@@
 ```{r}
-#| label: fig-coded
-#| fig-cap: Visualization to show correlation between number of Calls Coded (represented by the red line) and Homeless Death Counts (represented by the blue line)
+#| label: fig-total-delay
+#| fig-cap: Visualization to show the TTC service that has the highest number of delays in year 2023
 #| echo: false
 #| message: false
+#construct appropriate table for visualization
+service_type <- c("bus", "streetcar", "subway")
+num_delay <- c(nrow(bus_delay), nrow(streetcar_delay), nrow(subway_delay))
 
-combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
-                          show_col_types = FALSE)
+delay_count <- data.frame(service_type, num_delay)
 
-combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
-  geom_point() + geom_smooth() + 
-  geom_line(aes(x = Month, y = Calls_Coded_month/1000), color = "red") +
-  labs(
-    title = "Calls Coded & Homeless Death Count",
-    x = "Time (in Month)",
-    y = "Count"
-  )
-#still need to add a legend or label, but I haven't yep figured out how
+# Generate the bar plot
+ggplot(delay_count, aes(x = service_type, y = num_delay, fill = service_type)) +
+  geom_bar(stat = "identity") + 
+  labs(title = "Number of Delays by Service Type", x = "Service Type", y = "Number of Delays", fill = "Service Type") +  # Label the axes
+  theme_minimal()  # Apply a minimal theme
 ```
@@
 Present Results During Rush Hours & at Off-Peak Hours
+
 ```{r}
-#| label: fig-referred
-#| fig-cap: Visualization to show correlation between number of callers referred to shelter (represented by the purple line) and Homeless Death Counts (represented by the blue line)
 #| echo: false
@@
 
-combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
-                          show_col_types = FALSE)
+#filter out delays that occurred on weekdays
+weekday_bus_delay <- bus_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
+weekday_streetcar_delay <- streetcar_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
+weekday_subway_delay <- subway_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
 
-combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
-  geom_point() + geom_smooth() + 
-  geom_line(aes(x = Month, y = Referral_to_Shelter_month/100), color = "purple") +
-  labs(
-    title = "Callers Referred to Shelter & Homeless Death",
-    x = "Time (in Month)",
-    y = "Count"
-  )
-#still need to add a legend or label, but I haven't yep figured out how
+#filter out delays that occurred on weekends
+weekend_bus_delay <- bus_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+weekend_streetcar_delay <- streetcar_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+weekend_subway_delay <- subway_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+```
+
+```{r}
+#| label: fig-weekday
+#| fig-cap: Visualization to show the most delayed services on weekdays rush hours and off-peak hours
+#| echo: false
+#| message: false
+
+```
+
+```{r}
+#| label: fig-weekend
+#| fig-cap: Visualization to show the most delayed services on weekdays rush hours
+#| echo: false
+#| message: false
+
 ```
@@
 
-combined_data <- read_csv(here::here("outputs/data/combined_data.csv"),
-                          show_col_types = FALSE)
-
-combined_data|> ggplot(mapping = aes(x = Month, y = Death_Count)) + 
-  geom_point() + geom_smooth() + 
-  geom_line(aes(x = Month, y = Information_Homelessness_and_Prevention_month/100), color = "darkgreen") +
-  labs(
-    title = "Callers given Info on Homeless Prevention vs Homeless Death",
-    x = "Time (in Month)",
-    y = "Count"
-  )
-#still need to add a legend or label, but I haven't yep figured out how
 ```
@@
 - 1 table for all three services
+- separate into rush & off-peak hours
 

### 2024-09-22 outputs/paper/paper.qmd
moved data-filtering steps (initially in result section of paper) to cleaning script due to excess code chunks in paper (unorganized)

@@
 library(ggplot2)
+library(hms)
 
-bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus_delay.csv"))
-streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar_delay.csv"))
-subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway_delay.csv"))
+bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay.csv"))
+streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay.csv"))
+subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay.csv"))
 subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))
@@
   geom_bar(stat = "identity") + 
+  geom_text(aes(label = paste0(num_delay, " delayed")), 
+          vjust = 1, color = "white", size = 3.5, position = position_stack(vjust = 0.55)) +
   labs(title = "Number of Delays by Service Type", x = "Service Type", y = "Number of Delays", fill = "Service Type") +  # Label the axes
@@
 ## Most Delayed during Weekday
-Define Rush Hours
+Define Rush Hours:
+  Rush Hours: 
+  •	Morning Rush-Hour: Approximately 7:00 AM to 9:00 AM
+	•	Evening Rush-Hour: Approximately 4:00 PM to 7:00 PM
 
@@
 #| message: false
-
-#filter out delays that occurred on weekdays
-weekday_bus_delay <- bus_delay |>
-  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
-weekday_streetcar_delay <- streetcar_delay |>
-  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
-weekday_subway_delay <- subway_delay |>
-  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
-
-#filter out delays that occurred on weekends
-weekend_bus_delay <- bus_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
-weekend_streetcar_delay <- streetcar_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
-weekend_subway_delay <- subway_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
+weekday_bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay.csv"))
+weekday_streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay.csv"))
+weekday_subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay.csv"))
 ```
@@
 #| label: fig-weekday
-#| fig-cap: Visualization to show the most delayed services on weekdays rush hours and off-peak hours
+#| fig-cap: Visualization to show the total delayed time during weekdays in 2023
 #| echo: false
@@
 
-```
+service_type <- c("bus", "streetcar", "subway")
+sum_delay_time <- c(sum(weekday_bus_delay$min_delay), sum(weekday_streetcar_delay$min_delay), sum(weekday_subway_delay$min_delay))
 
-```{r}
-#| label: fig-weekend
-#| fig-cap: Visualization to show the most delayed services on weekdays rush hours
-#| echo: false
-#| message: false
+total_delay_time <- data.frame(service_type, as.numeric(sum_delay_time))
+
+# Generate the bar plot
+ggplot(total_delay_time, aes(x = service_type, y = sum_delay_time, fill = service_type)) +
+  geom_bar(stat = "identity") + 
+  # Add labels for total hours
+  geom_text(aes(label = paste0(round(sum_delay_time / 60, 2), " hours")), 
+            vjust = 1, color = "white", size = 3.5, position = position_stack(vjust = 0.55)) +  
+  labs(title = "Total Minutes Delayed in 2023 during Weekdays", 
+       x = "Service Type", y = "Total Delayed (min)", fill = "Service Type") +  # Label the axes
+  theme_minimal()
 
@@
 
-## Most Delayed during Weekends
+## Most Delayed during Rush-Hours & Off-Peak Hours
 
 ```{r}
-#| label: fig-informed
-#| fig-cap: Visualization to show correlation between number of callers given information on homelessness prevention (represented by the dark green line) and Homeless Death Counts (represented by the blue line)
+#| label: fig-rush-off
+#| fig-cap: Horizontal Barplot used to visualize the longest average delay time for different TTC services during Rush-Hours and Off-Peak Hours
 #| echo: false

### 2024-09-22 scripts/02-data_cleaning.R

@@ -10,6 +10,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(janitor)
+library(dplyr)
 library(readxl)
 library(readr)
 
@@ -60,12 +61,93 @@ subway_delay_code <- subway_delay_code |>
 #rbind the two datasets
 subway_delay_code <- rbind(subway_delay_code, temp_data)
 
+
+### Build Filterized Dataset Based on Mastercopy Delay Datasets ###
+
+## Weekday Delay
+#filter out delays that occurred on weekdays
+weekday_bus_delay <- bus_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
+weekday_streetcar_delay <- streetcar_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
+weekday_subway_delay <- subway_delay |>
+  filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
+
+# Weekend Delay
+weekend_bus_delay <- bus_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+weekend_streetcar_delay <- streetcar_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+weekend_subway_delay <- subway_delay |>
+  filter(day %in% c("Saturday", "Sunday"))
+
+## Rush-Hour Delay
+# Bus
+bus_delay_time <- bus_delay |>
+  mutate(time = as_hms(paste0(time, ":00"))) |>  # Convert 'time' to HH:MM:SS format
+  arrange(time)  # Sort 'time' column in ascending order
+
+bus_delay_rush <- bus_delay_time |>
+  filter(time >= as_hms("07:00:00") & time <= as_hms("19:00:00")) |>
+  filter(time < as_hms("09:00:00") | time > as_hms("16:00:00")) |>
+  mutate(rush_type = ifelse(time <= as_hms("09:00:00"), "morning", "evening"))
+
+bus_delay_off_peak <- bus_delay_time %>%
+  filter((time < as_hms("07:00:00")) | 
+           (time > as_hms("09:00:00") & time < as_hms("16:00:00")) | 
+           (time > as_hms("19:00:00")))
+# Streetcar
+streetcar_delay_time <- streetcar_delay |>
+  mutate(time = as_hms(paste0(time, ":00"))) |>  # Convert 'time' to HH:MM:SS format
[…]

### 2024-09-23 outputs/paper/paper.qmd
created rush-hour visualization and deleted excess files

@@
 ---
-title: "Choosing TTC x Provides the Most Time Stability during Rush Hours"
+title: "Choosing TTC Subway Service Provides the Most Time Stability during Weekday Rush-Hours"
 subtitle: "Analysis on the Delay Magnitude of different TTC Services"
@@
 date-format: long
-abstract: ""
+abstract: "This paper analyzes the magnitude of TTC delays in 2023 based on service types.
+TTC delay data for buses, streetcars, and subways are obtained through Open Data Toronto data portal and data is used to visualize delay magnitude during different days and hours throughout the year.
+We find that, although TTC streetcars offered the least number of delays in 2023, riding on TTC subways provides the least average time of delay during weekday rush-hour traffic.
+Our results suggest that, if the time required to arrive at one's office is the same for all TTC service options, choosing subway over other services could provide the most time stability."
 format: pdf
@@
 library(lubridate)
-library(zoo)
 library(knitr)
@@
 library(hms)
+library(patchwork)
 
@@
 #| message: false
+# Setup
 weekday_bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay.csv"))
@@
 sum_delay_time <- c(sum(weekday_bus_delay$min_delay), sum(weekday_streetcar_delay$min_delay), sum(weekday_subway_delay$min_delay))
+mean_delay_time <- c(mean(weekday_bus_delay$min_delay), mean(weekday_streetcar_delay$min_delay), mean(weekday_subway_delay$min_delay))
 
 total_delay_time <- data.frame(service_type, as.numeric(sum_delay_time))
+avg_delay_time <- data.frame(service_type, as.numeric(mean_delay_time))
 
 # Generate the bar plot
-ggplot(total_delay_time, aes(x = service_type, y = sum_delay_time, fill = service_type)) +
+sum_graph <- ggplot(total_delay_time, aes(x = service_type, y = sum_delay_time/60, fill = service_type)) +
   geom_bar(stat = "identity") + 
   # Add labels for total hours
-  geom_text(aes(label = paste0(round(sum_delay_time / 60, 2), " hours")), 
-            vjust = 1, color = "white", size = 3.5, position = position_stack(vjust = 0.55)) +  
-  labs(title = "Total Minutes Delayed in 2023 during Weekdays", 
-       x = "Service Type", y = "Total Delayed (min)", fill = "Service Type") +  # Label the axes
-  theme_minimal()
+  geom_text(aes(label = paste0(round(sum_delay_time / 60, 2), "hrs")), 
+            vjust = 1, color = "black", size = 1.5, lineheight = 0.8, position = position_stack(vjust = 0.6)) +  
+  labs(title = "Total Weekday Delay (hrs)", 
+       x = "Service Type", y = "Total Delayed (hrs)", fill = "Service Type") +  # Label the axes
+  theme_minimal() +
+  theme(legend.position = "none", axis.text.x = element_text(size = 8))
 
+avg_graph <- ggplot(avg_delay_time, aes(x = service_type, y = mean_delay_time, fill = service_type)) +
+  geom_bar(stat = "identity") + 
+  # Add labels for total hours
+  geom_text(aes(label = paste0(round(mean_delay_time, 1), " mins")), 
+            vjust = 1, color = "black", size = 2, position = position_stack(vjust = 0.6)) +  
+  labs(title = "Average Weekday Delay (min)", 
+       x = "Service Type", y = "Average Delay Time (min)", fill = "Service Type") +  # Label the axes
+  theme_minimal() +
+  theme(axis.text.x = element_text(size = 8))
+
+sum_graph + avg_graph
 ```
 
-## Most Delayed during Rush-Hours & Off-Peak Hours
+## Most Delayed during Rush-Hours
+```{r}
+#| echo: false
+#| message: false
+# Setup
+bus_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay_rush.csv"))
+streetcar_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay_rush.csv"))
+subway_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay_rush.csv"))
+```
 
@@
 #| label: fig-rush-off
-#| fig-cap: Horizontal Barplot used to visualize the longest average delay time for different TTC services during Rush-Hours and Off-Peak Hours
+#| fig-cap: Barplot used to visualize the longest average delay time for different TTC services during Rush-Hours and Off-Peak Hours
 #| echo: false
@@
 
+service_type <- c("bus", "streetcar", "subway")
+
+morning_bus_rush <- bus_delay_rush |> filter(rush_type == "morning")
+morning_streetcar_rush <- streetcar_delay_rush |> filter(rush_type == "morning")
+morning_subway_rush <- subway_delay_rush |> filter(rush_type == "morning")
+morning_mean_delay_rush <- c(mean(morning_bus_rush$min_delay), mean(morning_streetcar_rush$min_delay), mean(morning_subway_rush$min_delay))
+
+#build morning rush dataset
+morning_avg_delay_rush <- data.frame(service_type, as.numeric(morning_mean_delay_rush))
+
+evening_bus_rush <- bus_delay_rush |> filter(rush_type == "evening")
+evening_streetcar_rush <- streetcar_delay_rush |> filter(rush_type == "evening")
+evening_subway_rush <- subway_delay_rush |> filter(rush_type == "evening")
+evening_mean_delay_rush <- c(mean(evening_bus_rush$min_delay), mean(evening_streetcar_rush$min_delay), mean(evening_subway_rush$min_delay))
+
+#build evening rush dataset
+evening_avg_delay_rush <- data.frame(service_type, as.numeric(evening_mean_delay_rush))
+
+morning_rush <- ggplot(morning_avg_delay_rush, aes(x = service_type, y = morning_mean_delay_rush, fill = service_type)) + geom_bar(stat = "identity") + 
+  # Add labels for total hours
+  geom_text(aes(label = paste0(round(mean_delay_time, 1), " mins")), 
+            vjust = 1, color = "white", size = 2, position = position_stack(vjust = 0.6)) +  
+  labs(title = "Morning Rush Average Delay", 
+       x = "Service Type", y = "Average Delay Time (min)", fill = "Service Type") +  # Label the axes
+  theme_minimal() +
+  theme(legend.position = "none", axis.text.x = element_text(size = 8))
+
+evening_rush <- ggplot(evening_avg_delay_rush, aes(x = service_type, y = evening_mean_delay_rush, fill = service_type)) + geom_bar(stat = "identity") + 
+  # Add labels for total hours
+  geom_text(aes(label = paste0(round(mean_delay_time, 1), " mins")), 
+            vjust = 1, color = "white", size = 2, position = position_stack(vjust = 0.6)) +  
+  labs(title = "Evening Rush Average Weekday Delay", 
+       x = "Service Type", y = "Average Delay Time (min)", fill = "Service Type") +  # Label the axes
+  theme_minimal() +
+  theme(axis.text.x = element_text(size = 8))
+
+morning_rush + evening_rush
 ```
@@
 - separate into rush & off-peak hours
+```{r}
+
+```
 

### 2024-09-23 scripts/02-data_cleaning.R

@@ -73,17 +73,10 @@ weekday_streetcar_delay <- streetcar_delay |>
 weekday_subway_delay <- subway_delay |>
   filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
 
-# Weekend Delay
-weekend_bus_delay <- bus_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
-weekend_streetcar_delay <- streetcar_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
-weekend_subway_delay <- subway_delay |>
-  filter(day %in% c("Saturday", "Sunday"))
 
 ## Rush-Hour Delay
 # Bus
-bus_delay_time <- bus_delay |>
+bus_delay_time <- weekday_bus_delay |>
   mutate(time = as_hms(paste0(time, ":00"))) |>  # Convert 'time' to HH:MM:SS format
   arrange(time)  # Sort 'time' column in ascending order
 
@@ -92,12 +85,8 @@ bus_delay_rush <- bus_delay_time |>
   filter(time < as_hms("09:00:00") | time > as_hms("16:00:00")) |>
   mutate(rush_type = ifelse(time <= as_hms("09:00:00"), "morning", "evening"))
 
-bus_delay_off_peak <- bus_delay_time %>%
-  filter((time < as_hms("07:00:00")) | 
-           (time > as_hms("09:00:00") & time < as_hms("16:00:00")) | 
-           (time > as_hms("19:00:00")))
 # Streetcar
-streetcar_delay_time <- streetcar_delay |>
+streetcar_delay_time <- weekday_streetcar_delay |>
   mutate(time = as_hms(paste0(time, ":00"))) |>  # Convert 'time' to HH:MM:SS format
   arrange(time)  # Sort 'time' column in ascending order
 
@@ -106,13 +95,9 @@ streetcar_delay_rush <- streetcar_delay_time |>
   filter(time < as_hms("09:00:00") | time > as_hms("16:00:00")) |>
   mutate(rush_type = ifelse(time <= as_hms("09:00:00"), "morning", "evening"))
 
-streetcar_delay_off_peak <- streetcar_delay_time %>%
-  filter((time < as_hms("07:00:00")) | 
-           (time > as_hms("09:00:00") & time < as_hms("16:00:00")) | 
-           (time > as_hms("19:00:00")))
 
 # Subway
-subway_delay_time <- subway_delay |>
[…]

### 2024-09-23 outputs/paper/paper.qmd
fixed table styling

@@
 
-knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Direction", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
+knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay", "Min Gap", "Direction", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
 ```
@@
 
-knitr::kable(sample_streetcar_delay, col.names = c("Date", "Line", "Time", "Day", "Location", "Cause of Delay", "Min Delayed", "Min Gap", "Bound", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
+library(knitr)
+library(kableExtra)
+
+# Generate the table with automatic text wrapping for column names
+knitr::kable(sample_streetcar_delay, 
+             col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay", "Min Gap", "Bound", "Vehicle"), align = c("l","c", "c", "c", "c", "c", "c", "c", "c", "c")) 
 ```
@@
 
-knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Location", "Delay Code", "Min Delayed", "Min Gap", "Bound", "Line", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
+knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Station", "Delay Code", "Min Delayed", "Min Gap", "Bound", "Line", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
 ```

### 2024-09-23 outputs/paper/paper.qmd
fixed library message showing issue
completed all visualization in Result section and re-rendered paper.pdf

@@
 sample_streetcar_delay <- head(streetcar_delay, n = 3)
-
-library(knitr)
-library(kableExtra)
 
@@
 ```{r}
+#| label: tbl-reason
+#| tbl-cap: Top 5 Most Popular Reason for Delay
+#| echo: false
+#| message: false
 
+# Extract the top 5 most frequent 'incident' values for each dataset
+top_bus_incidents <- head(sort(table(bus_delay$incident), decreasing = TRUE), 5)
+top_streetcar_incidents <- head(sort(table(streetcar_delay$incident), decreasing = TRUE), 5)
+top_subway_incidents <- head(sort(table(subway_delay$code), decreasing = TRUE), 5)
+top_subway_codes <- names(top_subway_incidents)
+
+reason_delay <- data.frame(
+  names(top_bus_incidents), names(top_streetcar_incidents), top_subway_codes)
+
+# Create a new dataset 'reason_delay'
+reason_delay$top_subway_codes <- subway_delay_code$code_description[
+  match(reason_delay$top_subway_codes, subway_delay_code$code)
+]
+
+reason_delay <- reason_delay |> rename(bus = names.top_bus_incidents., streetcar = names.top_streetcar_incidents., subway = top_subway_codes)
+
+knitr::kable(reason_delay, col.names = c("Bus", "Streetcar", "Subway"), align = c("c", "c","c"))
 ```
 
-## Most Popular Delayed Vehicles
+## Most Popular Delayed Lines
 - 1 table for all three services
@@
 ```{r}
+#| label: tbl-lines
+#| tbl-cap: Most delayed Lines for each TTC Service
+#| echo: false
+#| message: false
 
+service_type <- c("bus", "streetcar", "subway")
+most_delay_line <- c(bus_delay$route[which.max(bus_delay$min_delay)], streetcar_delay$line[which.max(streetcar_delay$min_delay)], subway_delay$line[which.max(subway_delay$min_delay)])
+
+# create dataframe
+most_delay_data <- data.frame(service_type, most_delay_line)
+# rename SRT to proper TTC Line name
+most_delay_data$most_delay_line[most_delay_data$most_delay_line == "SRT"] <- "Line 3"
+
+#modify table aesthetics
+most_delay_transposed <- t(most_delay_data$most_delay_line)
+colnames(most_delay_transposed) <- most_delay_data$service_type
+
+# Display the transposed data using knitr::kable()
+knitr::kable(most_delay_transposed, col.names = colnames(most_delay_transposed), align = "c")
 ```

### 2024-09-23 outputs/paper/paper.qmd
Updated ReadMe Title and rerendered paper

@@
 
-## TTC Bus Delay
+## TTC Bus Delay {#sec-data-bus}
 
@@
 
-## TTC Streetcar Delay
+## TTC Streetcar Delay {#sec-data-streetcar}
 
@@
 
-## TTC Subway Delay
+## TTC Subway Delay {#sec-data-subway}
 
-### Delay Code
+### Delay Code {#sec-data-code}
 ```{r}

### 2024-09-24 outputs/paper/paper.qmd
constructed the backbone of the paper, started working on the intro seciton

--- 2024-09-23 outputs/paper/paper.qmd
@@
 # Introduction
+1) broader context to motivate
+- Public transportation is an important part of urban life, it is responsible for transporting millions of people around the city everyday.
+Choosing the correct TTC service offers a more stable timetable for office workers, student, and the elderly alike.
+
+2) some detail about what the paper is about; a clear gap that needs to be filled
+- TTC data has been publicly available since 2014, this paper seeks to use these open-sourced data to visualize how each type of TTC service performs through their delay time. 
+- TTC bus, streetcar, and subway delay data are all collected through @OpenDataToronto data portal.
+
+3) what was done
+- TTC bus, streetcar, and subway delay data in 2023 were extracted and compared to determine the TTC service type that offers the least amount of delay during weekday rush-hours.
+
+4) what was found
+- TTC subways has a history of offering the least average delay in 2023.
+
+5) why it is important
+- assist readers in choosing between different TTC services given that the distance between their residence and different types of TTC stations are the same.
 
@@
 # Data {#sec-data}
+Where the data comes from
 
-## TTC Bus Delay
+which datasets where chosen
+
+why choose 2023
+
+## TTC Bus Delay {#sec-data-bus}
+source of data (brief)
+structure of data & important attributes
+why are these attributes kept for analysis (measurement)
 
@@
 
-## TTC Streetcar Delay
+## TTC Streetcar Delay {#sec-data-streetcar}
+source of data (brief)
+structure of data & important attributes
+why are these attributes kept for analysis (measurement)
 
@@
 
-## TTC Subway Delay
+## TTC Subway Delay {#sec-data-subway}
+source of data (brief)
 
-### Delay Code
+### Delay Code {#sec-data-code}
+purpose of extra dataset
+how it is related to subway_delay
 ```{r}
@@
 ### Delay Data
+structure of data & important attributes
+why are these attributes kept for analysis (measurement)
+mention "code" column
 ```{r}
@@
 ```
-
-## TTC Delay Comparision
 

### 2024-09-24 outputs/paper/paper.qmd
finished intro section

@@
 # Introduction
-1) broader context to motivate
-- Public transportation is an important part of urban life, it is responsible for transporting millions of people around the city everyday.
-Choosing the correct TTC service offers a more stable timetable for office workers, student, and the elderly alike.
+Public transportation serves as the transportation backbone for city residents, facilitating the daily commute of millions of people.
+For office workers, students, and the elderly, selecting the appropriate TTC service is crucial in maintaining a reliable and efficient schedule.
+Despite the importance of this choice, there has been limited effort to evaluate the reliability of different TTC services to guide commuters’ choices.
 
-2) some detail about what the paper is about; a clear gap that needs to be filled
-- TTC data has been publicly available since 2014, this paper seeks to use these open-sourced data to visualize how each type of TTC service performs through their delay time. 
-- TTC bus, streetcar, and subway delay data are all collected through @OpenDataToronto data portal.
+Publicly accessible TTC delay data can be tracked back to 2014 on the @OpenDataToronto portal, yet a comprehensive visualization of how each type of service -- bus, streetcar, and subway -— performs in terms of delay time remains narrowly explored.
+This paper addresses this gap by using data from the TTC delay data to analyze and visualize delay patterns across different services.
 
-3) what was done
-- TTC bus, streetcar, and subway delay data in 2023 were extracted and compared to determine the TTC service type that offers the least amount of delay during weekday rush-hours.
+This paper specifically focuses on TTC bus, streetcar, and subway delay data from 2023, extracting and comparing them to identify which service type experienced the least amount of delay during weekday rush hours.
+Our findings indicate that, in 2023, the TTC subway system consistently demonstrated the shortest average delays.
 
-4) what was found
-- TTC subways has a history of offering the least average delay in 2023.
-
-5) why it is important
-- assist readers in choosing between different TTC services given that the distance between their residence and different types of TTC stations are the same.
+Understanding these patterns is crucial for helping commuters make informed choices about which TTC service to use, especially when they have equal access to different types of stations.
+By providing this analysis, the paper aim to assist readers in optimizing their travel choices and enhancing their overall commuting experience.
 

### 2024-09-24 outputs/paper/paper.qmd
cut out several unused columns in cleaned delay data and re-rendered

@@
 library(dplyr)
-library(lubridate)
 library(knitr)
@@
 Where the data comes from
+- All data used for and throughout the paper are extracted from Open Data Toronto Portal [@OpenDataToronto].
+TTC Bus Delay data [@Bus], TTC Streetcar Delay [@Streetcar], TTC Subway Delay [@Subway] and TTC Subway Delay Codes [@Subway] are retrieved to compare the delay magnitudes of different services.
+Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], and knitr [@knitr].
 
-which datasets where chosen
+For all delay dataset, only data in 2023 were chosen. There are three main reasons for this choice:
 
-why choose 2023
+1. 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns. 
+2. COVID-19 impacted TTC operating capacity and styles in other years close to 2024. 
+3. data from earlier years (before 2020) were structured very differently.
+
+Therefore only 2023 delay data were chosen in this paper.
 
@@
 
-knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay", "Min Gap", "Direction", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
+knitr::kable(sample_bus_delay, col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay"), align = c("l","c", "c", "c", "c","c", "c"))
 ```
@@
 knitr::kable(sample_streetcar_delay, 
-             col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay", "Min Gap", "Bound", "Vehicle"), align = c("l","c", "c", "c", "c", "c", "c", "c", "c", "c")) 
+             col.names = c("Date", "Line", "Time", "Day", "Station", "Cause of Delay", "Min Delay"), align = c("l","c", "c", "c", "c", "c", "c")) 
 ```
@@
 
-knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Station", "Delay Code", "Min Delayed", "Min Gap", "Bound", "Line", "Vehicle"), align = c("l","c", "c", "c", "c","c", "c", "c", "c","c"))
+knitr::kable(sample_subway_delay, col.names = c("Date", "Time", "Day", "Station", "Delay Code", "Min Delayed", "Line"), align = c("l","c", "c", "c", "c","c", "c"))
 ```

### 2024-09-24 outputs/paper/paper.qmd
editted code style in paper and re-rendered

@@
 #| message: false
-# load librarys and datasets
+# load libraries and datasets
 library(tidyverse)
@@
 # Introduction
+
 Public transportation serves as the transportation backbone for city  residents, facilitating the daily commute of millions of people.
 For office workers, students, and the elderly, selecting the appropriate TTC service is crucial in maintaining a reliable and efficient schedule.
-Despite the importance of this choice, there has been limited effort to evaluate the reliability of different TTC services to guide commuters’ choices.
+Despite the importance of this choice, there has been limited effort to evaluate the reliability of different TTC services to guide commuters’choices.
 
@@
 
-Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion.
-The Data section introduces the context of the datasets and examines all datasets and variables kept for analysis.
+Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion. The Data section introduces the context
+of the datasets and examines all datasets and variables kept for analysis.
 The Result section focuses on visualizing and presenting the data results in Data section.
@@
 # Data {#sec-data}
-Where the data comes from
-- All data used for and throughout the paper are extracted from Open Data Toronto Portal [@OpenDataToronto].
+
+Where the data comes from - All data used for and throughout the paper are extracted from Open Data Toronto Portal [@OpenDataToronto].
 TTC Bus Delay data [@Bus], TTC Streetcar Delay [@Streetcar], TTC Subway Delay [@Subway] and TTC Subway Delay Codes [@Subway] are retrieved to compare the delay magnitudes of different services.
 Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], and knitr [@knitr].
+\
+\
+For all delay dataset, only data in 2023 were chosen. There are three
+main reasons for this choice:
 
-For all delay dataset, only data in 2023 were chosen. There are three main reasons for this choice:
+**1.** 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns.
 
-1. 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns. 
-2. COVID-19 impacted TTC operating capacity and styles in other years close to 2024. 
-3. data from earlier years (before 2020) were structured very differently.
+**2.** COVID-19 impacted TTC operating capacity and styles in other years close to 2024.
+
+**3.** data from earlier years (before 2020) were structured very differently.
 
@@
 ## TTC Bus Delay {#sec-data-bus}
-source of data (brief)
-structure of data & important attributes
+
+source of data (brief) structure of data & important attributes
+
 why are these attributes kept for analysis (measurement)
@@
 ## TTC Streetcar Delay {#sec-data-streetcar}
-source of data (brief)
-structure of data & important attributes
+
+source of data (brief) structure of data & important attributes
+
 why are these attributes kept for analysis (measurement)
@@
 ## TTC Subway Delay {#sec-data-subway}
+
 source of data (brief)
@@
 ### Delay Code {#sec-data-code}
-purpose of extra dataset
-how it is related to subway_delay
+
+purpose of extra dataset how it is related to subway_delay
+
 ```{r}
@@
 ### Delay Data
+
 structure of data & important attributes
-why are these attributes kept for analysis (measurement)
-mention "code" column
+
+why are these attributes kept for analysis (measurement) mention "code" column
+
 ```{r}
@@
 
-# Results {#sec-result}
+# Results {#sec-results}
 
@@
 ## Most Delayed during Weekday
-Define Rush Hours:
-  Rush Hours: 
+
+Define Rush Hours: Rush Hours: 
 • Morning Rush-Hour: Approximately 7:00 AM to 9:00 AM 
@@
 ## Most Delayed during Rush-Hours
+
 ```{r}
@@
 ## Most Popular Delay Reasons
+
 -   1 table for all three services
 -   separate into rush & off-peak hours
+
 ```{r}
@@
 ## Most Popular Delayed Lines
+
 -   1 table for all three services
 -   separate into rush & off-peak hours
+
 ```{r}
@@
 
-## Correlation between Delay and Direction of Travel
-- Financial District
-- Midtown Direction
+## Importance of TTC Services
+- capacity of TTC daily TTC service capacities
+- rush-hour traffic
 
 ## TTC Service Amenities
+
 - The PATH (for financial district)
@@
 ## Errors in analysis, Limitations and Next Steps
+- only included 2023 data, should include more years for fairer inspection
+- unable to distinguish between on-time rate for different district
+- next step: explore GO services (GO trains)
 

### 2024-09-24 outputs/paper/paper.qmd
added conent to discussion & fixed figure position in rendered product

@@
 TTC Bus Delay data [@Bus], TTC Streetcar Delay [@Streetcar], TTC Subway Delay [@Subway] and TTC Subway Delay Codes [@Subway] are retrieved to compare the delay magnitudes of different services.
-Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], and knitr [@knitr].
+Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], kableExtra [@kableExtra], and knitr [@knitr].
 \
@@
 ```{r}
-#| label: fig-rush-off
+#| label: fig-rush
 #| fig-cap: Barplot used to visualize the longest average delay time for different TTC services during Rush-Hours and Off-Peak Hours
+#| fig-pos: "H"
 #| echo: false
@@
             vjust = 1, color = "white", size = 2, position = position_stack(vjust = 0.6)) +  
-  labs(title = "Morning Rush Average Delay", 
+  labs(title = "AM-Rush Average Delay", 
        x = "Service Type", y = "Average Delay Time (min)", fill = "Service Type") +  # Label the axes
@@
             vjust = 1, color = "white", size = 2, position = position_stack(vjust = 0.6)) +  
-  labs(title = "Evening Rush Average Weekday Delay", 
+  labs(title = "PM-Rush Average Delay", 
        x = "Service Type", y = "Average Delay Time (min)", fill = "Service Type") +  # Label the axes
@@
 
-knitr::kable(reason_delay, col.names = c("Bus", "Streetcar", "Subway"), align = c("c", "c","c"))
+knitr::kable(reason_delay, col.names = c("Bus", "Streetcar", "Subway"), align = c("c", "c", "c")) |>
+  kableExtra::kable_styling(position = "center")
 ```

### 2024-09-24 outputs/paper/paper.qmd
added to data section and reference, re-rendered paper

@@
 
-source of data (brief) structure of data & important attributes
+TTC bus delay data are collected and published by the Toronto Transit Commission [@Bus], with latest update in September of 2024.
+A total of 7 attributes of the data are kept for further evaluation: *date of recorded delay, bus route, scheduled time of arrival, scheduled day of the week (ex.
+Sunday, Monday, etc.), station, cause of delay,* and *total minutes delayed*.
+Additional data attributes, such as *vehicle number*, were excluded from the final dataset as they are less relevant to the focus of our analysis—weekday rush-hour delay patterns.
 
-why are these attributes kept for analysis (measurement)
+Please see @tbl-bus for an overview of the data.
 
 ```{r}
-#| label: tbl-bus-delay
+#| label: tbl-bus
 #| tbl-cap: Sample of Cleaned TTC Bus Delay Dataset
@@
 
-source of data (brief) structure of data & important attributes
+Streetcar delay data are also collected and published by the Toronto Transit Commission [@Bus].
+Latest update of this dataset occurred in September of 2024.
+A total of 7 attributes of the data are kept for further evaluation: *date of recorded delay, streetcar line, scheduled time of arrival, scheduled day of the week* (ex.
+Sunday, Monday, etc.)*, station, cause of delay,* and *total minutes delayed*.
+Additional data attributes, such as *time gap between current and next vehicle*, were excluded from the final dataset as they are less relevant to the focus of our analysis—weekday rush-hour delay patterns.
 
-why are these attributes kept for analysis (measurement)
+@tbl-streetcar contains an overview of streetcar delay data.
 
 ```{r}
-#| label: tbl-streetcar-delay
+#| label: tbl-streetcar
 #| tbl-cap: Sample of Cleaned TTC Streetcar Delay Dataset
@@
 
-source of data (brief)
+Publisher of data, update cycle (last updated date)
+Streetcar delay data are also collected and published by the Toronto Transit Commission [@Bus]. Latest update of this dataset occurred in September of 2024.
 
-### Delay Code {#sec-data-code}
+@tbl-subway contains an overview of subway delay data.
 
-purpose of extra dataset how it is related to subway_delay
+### Delay Data {#sec-subway-delay}
+
+A total of 7 attributes of the data are kept from the 2023 subway delay data file for further evaluation: *date of recorded delay, scheduled time of arrival, scheduled day of the week* (ex.
+Sunday, Monday, etc.)*, station, cause of delay, total minutes delayed,* and *subway line*.
+One potentially confusing aspect of the subway data is the *cause of delay* attribute, which is represented as codes in the dataset.
+
+An additional dataset containing the code descriptions -- subway delay code -- was extracted and cleaned for reference. Please see @tbl-code for a sample of subway incident codes.
 
 ```{r}
-#| label: tbl-subway-code
-#| tbl-cap: Sample of TTC Subway Delay Code
-#| echo: false
-#| message: false
-
-sample_subway_delay_code <- head(subway_delay_code, n = 3)
-
-knitr::kable(sample_subway_delay_code, col.names = c("Code", "Code Description"), align = c("c","c"))
-```
-
-### Delay Data
-
-structure of data & important attributes
-
-why are these attributes kept for analysis (measurement) mention "code" column
-
-```{r}
-#| label: tbl-subway-delay
+#| label: tbl-subway
 #| tbl-cap: Sample of Cleaned TTC Subway Delay Dataset
@@
 
+### Delay Code {#sec-data-code}
+
+The table below contains a cleaned dataset used to reference TTC subway delay causes.
+This delay code dataset contains two key attributes: *code*, and *code description*.
+Delay code data is used throughout this paper as an reference for the subway delay data discussed in the @sec-subway-delay.
+
+```{r}
+#| label: tbl-code
+#| tbl-cap: Sample of TTC Subway Delay Code
+#| fig-pos: "H"
+#| echo: false
+#| message: false
+
+sample_subway_delay_code <- head(subway_delay_code, n = 3)
+
+knitr::kable(sample_subway_delay_code, col.names = c("Code", "Code Description"), align = c("c","c"))
+```
+
 # Results {#sec-results}
@@
 #| fig-cap: Visualization to show the TTC service that has the highest number of delays in year 2023
+#| fig-pos: "H"
 #| echo: false
@@
 #| fig-cap: Visualization to show the total delayed time during weekdays in 2023
+#| fig-pos: "H"
 #| echo: false
@@
 #| tbl-cap: Top 5 Most Popular Reason for Delay
+#| fig-pos: "H"
 #| echo: false
@@
 #| tbl-cap: Most delayed Lines for each TTC Service
+#| fig-pos: "H"
 #| echo: false

### 2024-09-24 outputs/paper/paper.qmd
solved rendered figure title overlapping issue in section 3.2

@@
   geom_text(aes(label = paste0(round(sum_delay_time / 60, 2), "hrs")), 
-            vjust = 1, color = "black", size = 1.5, lineheight = 0.8, position = position_stack(vjust = 0.6)) +  
+            vjust = 1, color = "black", size = 1.6, lineheight = 0.8, position = position_stack(vjust = 0.6)) +  
   labs(title = "Total Weekday Delay (hrs)", 
@@
   theme_minimal() +
-  theme(legend.position = "none", axis.text.x = element_text(size = 8))
+  theme(legend.position = "none", axis.text.x = element_text(size = 8), 
+        plot.title = element_text(size = 11))
 
@@
   # Add labels for total hours
-  geom_text(aes(label = paste0(round(mean_delay_time, 1), " mins")), 
+  geom_text(aes(label = paste0(round(mean_delay_time, 1), "mins")), 
             vjust = 1, color = "black", size = 2, position = position_stack(vjust = 0.6)) +  
@@
   theme_minimal() +
-  theme(axis.text.x = element_text(size = 8))
+  theme(axis.text.x = element_text(size = 8), 
+        plot.title = element_text(size = 11))
 

### 2024-09-24 outputs/paper/paper.qmd
fixed position of tables and re-rendered

@@
 Data is cleaned and analyzed using the open source statistical programming language R [@citeR] and supporting packages tidyverse [@tidyverse], janitor [@janitor], dplyr [@dplyr], hms [@hms], ggplot2 [@ggplot2], patchwork [@patchwork], kableExtra [@kableExtra], and knitr [@knitr].
-\
-\
+
 For all delay dataset, only data in 2023 were chosen. There are three
@@
 
-**1.** 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns.
+>**1.** 2023 data is the most recent full-year data and offers the best reference for current (2024) delay patterns.
 
-**2.** COVID-19 impacted TTC operating capacity and styles in other years close to 2024.
+>**2.** COVID-19 impacted TTC operating capacity and styles in other years close to 2024.
 
-**3.** data from earlier years (before 2020) were structured very differently.
+>**3.** data from earlier years (before 2020) were structured very differently.
 
@@
 #| tbl-cap: Top 5 Most Popular Reason for Delay
-#| fig-pos: "H"
+#| table.position: "H"
 #| echo: false
@@
 knitr::kable(reason_delay, col.names = c("Bus", "Streetcar", "Subway"), align = c("c", "c", "c")) |>
-  kableExtra::kable_styling(position = "center")
+  kableExtra::kable_styling(position = "center", font_size = 10)
 ```
@@
 #| tbl-cap: Most delayed Lines for each TTC Service
-#| fig-pos: "H"
+#| table.position: "H"
 #| echo: false

### 2024-09-24 outputs/paper/paper.qmd
added to result section

@@
 # Results {#sec-results}
+The results section provides a comprehensive analysis of TTC service delays in 2023, with a focus on identifying patterns and factors contributing to these delays across bus, streetcar, and subway services.
+We begin by examining which service experienced the highest total number of delays, followed by an evaluation of delays specifically occurring on weekdays.
+Subsequently, we delve into delay patterns during rush hours to understand how peak travel times affect each service type.
+The section also explores the most common reasons for delays, highlighting issues unique to each service.
+Finally, we identify the routes most frequently affected by delays, offering insights into potential areas for targeted improvements.
+This multi-faceted analysis aims to provide a clearer picture of how delays impact TTC services and which areas may require attention for enhancing service reliability.
 
 ## Highest Number of Total Delayed
+
+@fig-total-delay shows that the TTC bus service experienced the highest number of total delays in 2023, with 56,207 delays recorded, followed by the subway service with 22,949 delays, and the streetcar service with 14,413 delays.
+However, it is important to note that while buses are the most frequently delayed service, the number of delays doesn’t necessarily reflect the overall magnitude of delay experienced by passengers.
+The next section -- @sec-weekday -- will dive deeper into the average delay time experienced by passengers, specifically during rush-hours on weekdays.
 
@@
 
-## Most Delayed during Weekday
+## Most Delayed during Weekday {#sec-weekday}
 
-Define Rush Hours: Rush Hours: 
-• Morning Rush-Hour: Approximately 7:00 AM to 9:00 AM 
-• Evening Rush-Hour: Approximately 4:00 PM to 7:00 PM
-
-Present Results During Rush Hours & at Off-Peak Hours
+The figure below, @fig-weekday, reveals that the total weekday delay in hours was highest for buses, reaching approximately 18971.37 hours, followed by streetcars with 3736.2 hours, and finally, subways with 1136.32 hours.
+When examining the average delay per incident, buses experienced the longest average delay time at 20.3 minutes, streetcars averaged 15.6 minutes, and subways had the shortest average delay of just 3 minutes per incident.
+These two visualizations complement each other to indicate that TTC subway consistently exhibited the least delay on weekdays, making it the most reliable option for weekday travel.
 
@@
 
-## Most Delayed during Rush-Hours
+### Most Delayed during Rush-Hours
 

### 2024-09-24 outputs/paper/paper.qmd
refined additional result section

@@
 This paper specifically focuses on TTC bus, streetcar, and subway delay data from 2023, extracting and comparing them to identify which service type experienced the least amount of delay during weekday rush hours.
+Throughout the paper, *rush-hour* will refers to the periods from 7:00 AM to 9:00 AM (morning rush) and 4:00 PM to 7:00 PM (evening rush).
 Our findings indicate that, in 2023, the TTC subway system consistently demonstrated the shortest average delays.
@@
 TTC bus delay data are collected and published by the Toronto Transit Commission [@Bus], with latest update in September of 2024.
-A total of 7 attributes of the data are kept for further evaluation: *date of recorded delay, bus route, scheduled time of arrival, scheduled day of the week (ex.
-Sunday, Monday, etc.), station, cause of delay,* and *total minutes delayed*.
+A total of 7 attributes of the data are kept for further evaluation: *date of recorded delay, bus route, scheduled time of arrival, scheduled day of the week* (ex.
+Sunday, Monday, etc.)*, station, cause of delay,* and *total minutes delayed*.
 Additional data attributes, such as *vehicle number*, were excluded from the final dataset as they are less relevant to the focus of our analysis—weekday rush-hour delay patterns.
@@
 
+Rush hour, as defined in this paper, refers to the periods from 7:00 AM to 9:00 AM in the morning and 4:00 PM to 7:00 PM in the evening.
+@fig-rush illustrates that during these periods, buses experienced the longest average delay times, with both the morning and evening rush periods showing an average delay of approximately 20.3 minutes.
+Streetcars followed, with an average delay of 15.6 minutes for both morning and evening rush hours, while subways had the shortest average delays, recording just 3 minutes in both periods.
+These results highlight that the subway consistently offers the most reliable service during rush-hour traffic compared to buses and streetcars.
+
 ```{r}
@@
 
--   1 table for all three services
--   separate into rush & off-peak hours
+@tbl-reason presents the top 5 most frequent reasons for delays across the three TTC services.
+For buses, the most common causes included mechanical issues and security, while streetcars experienced delays primarily due to operations and mechanical problems.
+For subways, disorderly patrons and passenger assistance alarms were among the top delay causes.
+Understanding these reasons is essential for addressing service inefficiencies.
 
@@
 
--   1 table for all three services
--   separate into rush & off-peak hours
+@tbl-lines highlights the lines most prone to delays for each TTC service.
+The bus route 121, the streetcar line 506, and subway Line 3 (SRT line) were identified as the most frequently delayed routes.
+These insights indicate potential areas for service improvement, especially during peak travel times.
 

### 2024-09-24 scripts/01-download_data.R
change directory for cleaned data and updated test scripts

@@ -27,19 +27,19 @@ bus_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/bus_delay_2023.
 ### Streetcar Data ###
 
 download.file("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/b68cb71b-44a7-4394-97e2-5d2f41462a5d/resource/472d838d-e41a-4616-a11b-585d26d59777/download/ttc-streetcar-delay-data-2023.xlsx",
-              "inputs/data/downloaded_xlsx_format/streetcar_delay_2023.xlsx", mode = "wb")
-streetcar_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/streetcar_delay_2023.xlsx")
+              "inputs/raw_data/streetcar_delay_2023.xlsx", mode = "wb")
+streetcar_delay_2023 <- read_excel("inputs/raw_data/streetcar_delay_2023.xlsx")
 
 
 
 ### Subway Data ###
 
 download.file("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/2fbec48b-33d9-4897-a572-96c9f002d66a/download/ttc-subway-delay-data-2023.xlsx",
-              "inputs/data/downloaded_xlsx_format/subway_delay_2023.xlsx", mode = "wb")
-subway_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/subway_delay_2023.xlsx")
+              "inputs/raw_data/subway_delay_2023.xlsx", mode = "wb")
+subway_delay_2023 <- read_excel("inputs/raw_data/subway_delay_2023.xlsx")
 
 ## Subway Delay Code ##
 
 download.file("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/996cfe8d-fb35-40ce-b569-698d51fc683b/resource/3900e649-f31e-4b79-9f20-4731bbfd94f7/download/ttc-subway-delay-codes.xlsx", 
-              "inputs/data/downloaded_xlsx_format/subway_delay_code.xlsx", mode = "wb")
-subway_delay_code <- read_excel("inputs/data/downloaded_xlsx_format/subway_delay_code.xlsx")
+              "inputs/raw_data/subway_delay_code.xlsx", mode = "wb")
+subway_delay_code <- read_excel("inputs/raw_data/subway_delay_code.xlsx")

### 2024-09-24 scripts/03-test_data.R

@@ -11,10 +11,15 @@
 library(testthat)
 
 #### Load data ####
-bus_delay <- read.csv("inputs/data/cleaned_data/bus_delay.csv")
-streetcar_delay <- read.csv("inputs/data/cleaned_data/streetcar_delay.csv")
-subway_delay <- read.csv("inputs/data/cleaned_data/subway_delay.csv")
-subway_delay_code <- read.csv("inputs/data/cleaned_data/subway_delay_code.csv")
+bus_delay <- read.csv("outputs/cleaned_data/bus/bus_delay.csv")
+weekday_bus_delay <- read.csv("outputs/cleaned_data/bus/weekday_bus_delay.csv")
+
+streetcar_delay <- read.csv("outputs/cleaned_data/streetcar/streetcar_delay.csv")
+weekday_streetcar_delay <- read.csv("outputs/cleaned_data/streetcar/weekday_streetcar_delay.csv")
+
+subway_delay <- read.csv("outputs/cleaned_data/subway/subway_delay.csv")
+weekday_subway_delay <- read.csv("outputs/cleaned_data/subway/weekday_subway_delay.csv")
+subway_delay_code <- read.csv("outputs/cleaned_data/subway_delay_code.csv")
 
 #### Test data ####
 
@@ -55,7 +60,18 @@ test_that("Check that dates are within 2023", {
               info = "All dates should be between 2023-01-01 and 2023-12-31")
 })
 
+# Test to check there are no NA values in the weekday dataset
+test_that("No NA values in date, time, and min_delay columns", {
+  expect_true(all(!is.na(weekday_bus_delay$date)))
+  expect_true(all(!is.na(weekday_bus_delay$time)))
+  expect_true(all(!is.na(weekday_bus_delay$min_delay)))
+})
 
+# Test to check that the 'day' column only contains weekdays
+test_that("Day column only contains Monday to Friday", {
+  valid_days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
+  expect_true(all(weekday_bus_delay$day %in% valid_days))
+})
 
 
 ### Streetcar Data ###
@@ -95,7 +111,19 @@ test_that("Check that dates are within 2023", {
               info = "All dates should be between 2023-01-01 and 2023-12-31")
 })
 
+# Test for weekday
+# Check for no NA values in date, time, and min_delay columns
+test_that("No NA values in date, time, and min_delay columns for weekday_streetcar_delay", {
+  expect_true(all(!is.na(weekday_streetcar_delay$date)))
[…]

### 2024-09-24 scripts/03-test_data.R
added aproppriate data loading function

@@ -13,12 +13,15 @@ library(testthat)
 #### Load data ####
 bus_delay <- read.csv("outputs/cleaned_data/bus/bus_delay.csv")
 weekday_bus_delay <- read.csv("outputs/cleaned_data/bus/weekday_bus_delay.csv")
+bus_delay_rush <- read.csv("outputs/cleaned_data/bus/bus_delay_rush.csv")
 
 streetcar_delay <- read.csv("outputs/cleaned_data/streetcar/streetcar_delay.csv")
 weekday_streetcar_delay <- read.csv("outputs/cleaned_data/streetcar/weekday_streetcar_delay.csv")
+streetcar_delay_rush <- read.csv("outputs/cleaned_data/bus/streetcar_delay_rush.csv")
 
 subway_delay <- read.csv("outputs/cleaned_data/subway/subway_delay.csv")
 weekday_subway_delay <- read.csv("outputs/cleaned_data/subway/weekday_subway_delay.csv")
+subway_delay_rush <- read.csv("outputs/cleaned_data/bus/subway_delay_rush.csv")
 subway_delay_code <- read.csv("outputs/cleaned_data/subway_delay_code.csv")
 
 #### Test data ####

### 2024-09-24 scripts/02-data_cleaning.R
updated saving directory in download script

@@ -13,13 +13,14 @@ library(janitor)
 library(dplyr)
 library(readxl)
 library(readr)
+library(hms)
 
 
 #### Read data ####
-bus_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/bus_delay_2023.xlsx")
-streetcar_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/streetcar_delay_2023.xlsx")
-subway_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/subway_delay_2023.xlsx")
-subway_delay_code <- read_excel("inputs/data/downloaded_xlsx_format/subway_delay_code.xlsx")
+bus_delay_2023 <- read_excel("inputs/raw_data/bus_delay_2023.xlsx")
+streetcar_delay_2023 <- read_excel("inputs/raw_data/streetcar_delay_2023.xlsx")
+subway_delay_2023 <- read_excel("inputs/raw_data/subway_delay_2023.xlsx")
+subway_delay_code <- read_excel("inputs/raw_data/subway_delay_code.xlsx")
 
 #### Clean data ####
 
@@ -27,18 +28,21 @@ subway_delay_code <- read_excel("inputs/data/downloaded_xlsx_format/subway_delay
 bus_delay <-
   bus_delay_2023 |>
   janitor::clean_names() |>
-  mutate(date = as.Date(date, format = "%Y-%m-%d"))
+  mutate(date = as.Date(date, format = "%Y-%m-%d")) |>
+  select(date, route, time, day, location, incident, min_delay)
 
 streetcar_delay <-
   streetcar_delay_2023 |>
   janitor::clean_names() |>
-  mutate(date = as.Date(date, format = "%Y-%m-%d"))
+  mutate(date = as.Date(date, format = "%Y-%m-%d")) |>
+  select(date, line, time, day, location, incident, min_delay)
 
 subway_delay <-
   subway_delay_2023 |>
   janitor::clean_names() |>
   mutate(date = as.Date(date, format = "%Y-%m-%d")) |>
-  rename(location = station)
+  rename(location = station) |>
+  select(date, time, day, location, code, min_delay, line)
 
 
 ### Subway Delay Code ###
@@ -109,20 +113,20 @@ subway_delay_rush <- subway_delay_time |>
 
 
 #### Save data ####
[…]

### 2024-09-24 scripts/01-download_data.R
updated old directory for bus_delay

@@ -19,8 +19,8 @@ library(readr)
 ### Bus Data ###
 
 download.file("https://ckan0.cf.opendata.inter.prod-toronto.ca/dataset/e271cdae-8788-4980-96ce-6a5c95bc6618/resource/10802a64-9ac0-4f2e-9538-04800a399d1e/download/ttc-bus-delay-data-2023.xlsx",
-              "inputs/data/downloaded_xlsx_format/bus_delay_2023.xlsx", mode = "wb")
-bus_delay_2023 <- read_excel("inputs/data/downloaded_xlsx_format/bus_delay_2023.xlsx")
+              "inputs/raw_data/bus_delay_2023.xlsx", mode = "wb")
+bus_delay_2023 <- read_excel("inputs/raw_data/bus_delay_2023.xlsx")
 
 
 

### 2024-09-26 outputs/paper/paper.qmd
leave addiional discussion section as bullet points
finished discussion and added conlusion to paper, re-rendered

@@
 
-bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay.csv"))
-streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay.csv"))
-subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay.csv"))
-subway_delay_code <- read.csv(here::here("inputs/data/cleaned_data/subway_delay_code.csv"))
+bus_delay <- read.csv(here::here("outputs/cleaned_data/bus/bus_delay.csv"))
+streetcar_delay <- read.csv(here::here("outputs/cleaned_data/streetcar/streetcar_delay.csv"))
+subway_delay <- read.csv(here::here("outputs/cleaned_data/subway/subway_delay.csv"))
+subway_delay_code <- read.csv(here::here("outputs/cleaned_data/subway_delay_code.csv"))
 
@@
 This paper specifically focuses on TTC bus, streetcar, and subway delay data from 2023, extracting and comparing them to identify which service type experienced the least amount of delay during weekday rush hours.
-Throughout the paper, *rush-hour* will refers to the periods from 7:00 AM to 9:00 AM (morning rush) and 4:00 PM to 7:00 PM (evening rush).
+Throughout the paper, *rush-hour* refers to the periods from 7:00 AM to 9:00 AM (morning rush) and 4:00 PM to 7:00 PM (evening rush).
 Our findings indicate that, in 2023, the TTC subway system consistently demonstrated the shortest average delays.
@@
 Understanding these patterns is crucial for helping commuters make informed choices about which TTC service to use, especially when they have equal access to different types of stations.
-By providing this analysis, the paper aim to assist readers in optimizing their travel choices and enhancing their overall commuting experience.
+By providing this analysis, the paper aims to assist readers in optimizing their travel choices and enhancing their overall commuting experience.
 
-Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion. The Data section introduces the context
-of the datasets and examines all datasets and variables kept for analysis.
+Analyses and findings in this paper are structured into several sections: @sec-data -- Data, @sec-results -- Results, and @sec-discussion -- Discussion.
+The Data section introduces the context of the datasets and examines all datasets and variables kept for analysis.
 The Result section focuses on visualizing and presenting the data results in Data section.
@@
 
-For all delay dataset, only data in 2023 were chosen. There are three
-main reasons for this choice:
+For all delay datasets, only data in 2023 was chosen. There are three main reasons for this choice:
 
@@
 @fig-total-delay shows that the TTC bus service experienced the highest number of total delays in 2023, with 56,207 delays recorded, followed by the subway service with 22,949 delays, and the streetcar service with 14,413 delays.
-However, it is important to note that while buses are the most frequently delayed service, the number of delays doesn’t necessarily reflect the overall magnitude of delay experienced by passengers.
+However, it is important to note that while buses are the most frequently delayed service, the number of delays does not necessarily reflect the overall magnitude of delay experienced by passengers.
 The next section -- @sec-weekday -- will dive deeper into the average delay time experienced by passengers, specifically during rush-hours on weekdays.
@@
 # Setup
-weekday_bus_delay <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay.csv"))
-weekday_streetcar_delay <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay.csv"))
-weekday_subway_delay <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay.csv"))
+weekday_bus_delay <- read.csv(here::here("outputs/cleaned_data/bus/bus_delay.csv"))
+weekday_streetcar_delay <- read.csv(here::here("outputs/cleaned_data/streetcar/streetcar_delay.csv"))
+weekday_subway_delay <- read.csv(here::here("outputs/cleaned_data/subway/subway_delay.csv"))
 ```
@@
 # Setup
-bus_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/bus/bus_delay_rush.csv"))
-streetcar_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/streetcar/streetcar_delay_rush.csv"))
-subway_delay_rush <- read.csv(here::here("inputs/data/cleaned_data/subway/subway_delay_rush.csv"))
+bus_delay_rush <- read.csv(here::here("outputs/cleaned_data/bus/bus_delay_rush.csv"))
+streetcar_delay_rush <- read.csv(here::here("outputs/cleaned_data/streetcar/streetcar_delay_rush.csv"))
+subway_delay_rush <- read.csv(here::here("outputs/cleaned_data/subway/subway_delay_rush.csv"))
 ```
@@
 ## TTC Service Prices {#sec-price}
--   one-fare
+All TTC services, whether you’re taking a bus, streetcar, or subway, operate under a one-fare system, meaning the financial cost of using any TTC service is identical.
+This pricing structure simplifies travel across the Greater Toronto Area (GTA), as passengers don’t need to worry about different costs when switching between services.
+The one-fare policy makes it easier for commuters to choose the most convenient transportation option without concern for cost differences, promoting greater accessibility and flexibility for daily travel across the city.
 
 ## Importance of TTC Services
-- capacity of TTC daily TTC service capacities
-- rush-hour traffic
+TTC services play a vital role in helping individuals navigate the GTA, providing an essential mode of transportation for millions of residents and visitors.
+According to the TTC’s official website [@ttc], these services have transported over 31 billion customers since their initial construction, highlighting the system’s critical role in supporting Toronto’s urban mobility.
+Despite the impact of the COVID-19 pandemic, the TTC has steadily regained ridership levels, currently operating at around 70% of its pre-pandemic capacity on weekdays and 75% to 90% on weekends.
+This re-bounce highlights the TTC’s importance in serving the city’s transportation needs, even as it continues to adjust to challenges after the pandemic.
 
@@
 
-- The PATH (for financial district)
-- AC in Summer & Winter
+One of the unique amenities of TTC services, particularly for subway passengers, is their integration with The PATH —- a vast underground pedestrian network that connects many downtown subway stations with offices and buildings in Toronto’s financial district.
+For commuters who work in buildings linked to The PATH, using subways or streetcars with direct PATH connections can provide a more convenient and sheltered travel experience, especially during extreme weather.
+Additionally, The PATH offers air conditioning during the summer and heating in the winter, ensuring a comfortable environment for passengers, making it a favorable choice over buses that lack such amenities.
+This integration highlights the advantages of certain TTC services, particularly for those commuting to downtown workplaces.
 
 ## Errors in analysis, Limitations and Next Steps
-- only included 2023 data, should include more years for fairer inspection
-- unable to distinguish between on-time rate for different district
-- next step: explore GO services (GO trains)
+The analysis presented in this paper is subject to certain limitations.
+Firstly, only data from 2023 was included, which may not provide a complete representation of TTC service performance over time.
+Including data from multiple years would offer a more comprehensive and fair evaluation of trends and patterns.
+Secondly, the study was unable to distinguish between on-time rates for different districts within Toronto, which limits the understanding of how delays vary across the different districts.
+As a next step, expanding the scope of the analysis to explore GO services, such as GO Trains, would provide a broader perspective on regional transit and allow for a comparison of how different transportation systems operate within and beyond the city.
 
 # Conclusion {#sec-conclusion}
+This paper has analyzed the delay patterns of TTC bus, streetcar, and subway services in 2023, revealing significant insights into their performance during weekday and rush-hour periods.
+The findings demonstrate that TTC buses experienced the highest number of delays, whereas subways consistently offered the least delay in terms of both total hours and average delay time.
+These results underscore the reliability of the subway as the most time-efficient option for commuters, especially during weekday rush hours.
+Understanding these patterns is essential for making informed transportation choices, and future research could further expand this analysis by incorporating multiple years of data and exploring other regional transit services such as GO Trains.
+This evaluation of TTC services will contribute to improving urban mobility and ensuring that Toronto’s public transportation system remains effective and reliable for all commuters.
 

### 2024-09-26 outputs/paper/paper.qmd
updated wording of abstract and intro

@@
 abstract: "This paper analyzes the magnitude of TTC delays in 2023 based on service types.
-TTC delay data for buses, streetcars, and subways are obtained through Open Data Toronto data portal and data is used to visualize delay magnitude during different days and hours throughout the year.
+Delay data for buses, streetcars, and subways are obtained through Open Data Toronto and the data collected are used to visualize delay magnitude during different days and hours throughout the year.
 We find that, although TTC streetcars offered the least number of delays in 2023, riding on TTC subways provides the least average time of delay during weekday rush-hour traffic.
@@
 
-Public transportation serves as the transportation backbone for city  residents, facilitating the daily commute of millions of people.
+Public transits serves as the transportation backbone for city  residents, facilitating the daily commute of millions of people.
 For office workers, students, and the elderly, selecting the appropriate TTC service is crucial in maintaining a reliable and efficient schedule.
